Praise the people or praise the place? Upper tail human capital in electrifying Sweden

### 2024-01-22 index.qmd
add presi

    - It made individuals richer 💰🔋
    - It lowered inequality 📉🤝
    - It allowed structural change on the labor market 🔧💼

@@

![](assets/vem_cover.png){fig-align="center" width=100%}

@@

## Vem är Vem? Example page

@@

We tell the system "You are an expert in Swedish family history"

Then we ask it to translate the Swedish text to English

@@

def translate_and_structure_text(swedish_text):
    try:
        # Translate the Swedish text to English
        translate_prompt = f"Translate the following abbreviated Swedish biography to English: {swedish_text}"
        translation_response = openai.ChatCompletion.create(
            model="gpt-3.5-turbo",
            messages=[
                {"role": "system", "content": "You are an expert on Swedish family history."},
                {"role": "user", "content": f"{translate_prompt}"}
            ]
        )

## 2. Structure 🤖

![](assets/schema.png)

## 2. Structure 🤖

We tell the system "You are an expert in Swedish family history and the Schema.org/Person format"

We provide it the original Swedish and English translation, and ask it to structure the biography in Schema.org/Person format as a JSON object

@@

structure_prompt = f"Given the original Swedish biography: {swedish_text}\nAnd its English translation: {english_text}, Structure the biography in Schema.org/Person format as a JSON object. Include dates wherever possible. Only provide a  RFC8259 compliant JSON response."

structure_response = openai.ChatCompletion.create(
    model="gpt-3.5-turbo",
    messages=[
        {"role": "system", "content": "You are an expert on Swedish family history and the Schema.org/Person format."},
        {"role": "user", "content": f"{structure_prompt}"}
    ]
)
```

## 3. Augment 🤖

### Using OpenAI's GPT-4 chat API:

We create python code to geocode any addresses in the biography (in this example), as well as flagging the highest level of education, and extract any mention of patents or inventions in the biography.

def geocode_birthplace(birthplace, geolocator, geocoded_cache):
    if isinstance(birthplace, dict):
        if "name" in birthplace:
            birthplace_name = birthplace["name"]
        elif "address" in birthplace and "addressLocality" in birthplace["address"]:
            birthplace_name = birthplace["address"]["addressLocality"]
        else:
            print(f"Invalid birthplace data: {birthplace}")
            return None
    elif isinstance(birthplace, str):
        birthplace_name = birthplace
    else:
        print(f"Invalid birthplace data: {birthplace}")
        return None

    if birthplace_name not in geocoded_cache:
        try:
            location = geolocator.geocode(birthplace_name)
            if location:
                geocoded_cache[birthplace_name] = {
                    "latitude": location.latitude,
                    "longitude": location.longitude
                }
            time.sleep(1)  # Add a 1-second delay between geocoding requests
        except Exception as e:
            print(f"Error geocoding birthplace {birthplace_name}: {e}")

    return geocoded_cache.get(birthplace_name)
```

## Output

{
    "original": "Lund, Karl Gustaf, överingenjör, Varberg, f. 22/7/93 i Hille, Gävleb. 1., av brukstj :m. Ferdinand L. o. Maria Andersson. G. 36 m. Sigrid Johansson. Barn: Ingvar f. 38, Lennart 42. — Ex. v. bergssk. i Filipstad 17, spec:-stud. v. KTH (B) 20-22, stud. v. me-tallogr. inst. o. Sthlms högsk. 21-22. Kemist v. Strömsnäs Järnverks A-B, Degerfors, 18-20, metallurg o. kemist v. Westinghouse Electric & Manuf. Co., East Pittsburgh, Pa, USA, 23-26 o. 28-29, chefsmetallurg v. Laclede Steel^Co., Ahon, 111., USA, 27, hytt-o. stålv.ing. v. A-B Iggesunds Bruk 29-31, platschef v. Gunnebo Bruks Nya A-B, Varbergsverket, sed. 31. Led. av drätselkamm., v. ordf. v. ekonomi-avd., suppl. i styr. f. elverket, huv:-man i Varbergs Sparbank, arb :giv. repr. i länsarb:ndns kretsråd, led. av styr. f. Varbergs luftsk ifören., sekr. i Varbergs högerfören., ordfs i järnv. sjukkassa o. Plant :sällsk. Småfågl. Vänner. Res. t. Tyskl. 21, 2^ 23, 30 o. 36, Danm., Tjeckoslov. 21, 22, 23, Österr. 21, USA 23-29. Skr.: Some fundamental factors for obtaining sharp thermal curves (Träns. Am. Soc. for Steel Treating, tills. m. C. Benedicks o. W. H. Dearden 25), Nutida fabrikation av sågblad, sågklingor o. maskinknivar (Trävaruind. 31). Hob-bies: jakt o. fiske.",
    "translated": "Lund, Karl Gustaf, chief engineer, Varberg, born on July 22, 1893, in Hille, Gävleborgs län, Sweden, son of factory worker Ferdinand L. and Maria Andersson. Married in 1936 to Sigrid Johansson. Children: Ingvar (born 1938), Lennart (born 1942). — Graduated from Bergsskolan in Filipstad in 1917, specialized studies at the Royal Institute of Technology (KTH) from 1920 to 1922, studied at the Institute of Metallurgy and Stockholm University in 1921-1922. Chemist at Strömsnäs Järnverks A-B in Degerfors from 1918 to 1920.",
    "structured": {
            "@context": "http://schema.org",
            "@type": "Person",
            "name": "Karl Gustaf Lund",
            "givenName": "Karl Gustaf",
            "familyName": "Lund",
            "gender": "Male",
            "birthDate": "1893-07-22",
            "birthPlace": {
                "@type": "Place",
                "address": {
                "@type": "PostalAddress",
                "addressLocality": "Hille",
                "addressRegion": "Gävleborgs län",
                "addressCountry": "Sweden"
                }
            },
            "nationality": "Swedish",
            "parent": [
                {
                "@type": "Person",
                "name": "Ferdinand L.",
                "jobTitle": "Factory Worker"
                },
                {
                "@type": "Person",
                "name": "Maria Andersson"
                }
            ],
            "spouse": {
                "@type": "Person",
                "name": "Sigrid Johansson",
                "marriageDate": "1936"
            },
            "children": [
                {
                "@type": "Person",
                "name": "Ingvar",
                "birthDate": "1938"
                },
                {
                "@type": "Person",
                "name": "Lennart",
                "birthDate": "1942"
                }
            ],
            "alumniOf": [
                {
                "@type": "EducationalOrganization",
                "name": "Bergsskolan in Filipstad",
                "graduationDate": "1917"
                },
                {
                "@type": "EducationalOrganization",
                "name": "Royal Institute of Technology (KTH)",
                "attendancePeriod": "1920-1922"
                },
                {
                "@type": "EducationalOrganization",
                "name": "Institute of Metallurgy and Stockholm University",
                "attendancePeriod": "1921-1922"
                }
            ],
            "worksFor": [
                {
                "@type": "Organization",
                "name": "Strömsnäs Järnverks A-B",
                "jobTitle": "Chemist",
                "employmentPeriod": "1918-1920"
                },
                {
                "@type": "Organization",
                "name": "Westinghouse Electric & Manufacturing Co.",
                "jobTitle": "Metallurgist and Chemist",
                "employmentPeriod": "1923-1926, 1928-1929",
                "location": "East Pittsburgh, PA, USA"
                },
                {
                "@type": "Organization",
                "name": "Laclede Steel Co.",
                "jobTitle": "Chief Metallurgist",
                "employmentPeriod": "1927",
                "location": "Alton, Illinois, USA"
                },
                {
                "@type": "Organization",
                "name": "A-B Iggesunds Bruk",
                "jobTitle": "Furnace and Steelworks Engineer",
                "employmentPeriod": "1929-1931"
                },
                {
                "@type": "Organization",
                "name": "Gunnebo Bruks Nya A-B, Varbergsverket",
                "jobTitle": "Site Manager",
                "employmentPeriod": "1931-present"
                }
            ],
            "memberOf": [
                {
                "@type": "Organization",
                "name": "Municipal Executive Committee"
                },
                {
                "@type": "Organization",
                "name": "Varbergs Sparbank",
                "position": "Chairman"
                },
                {
                "@type": "Organization",
                "name": "Varbergs Luftskyddsförening",
                "position": "Member of the Board"
                },
                {
                "@type": "Organization",
                "name": "Varbergs Högerförening",
                "position": "Secretary"
                }
            ],
            "hasOccupation": {
                "@type": "Occupation",
                "name": "Chief Engineer"
            },
            "address": {
                "@type": "PostalAddress",
                "addressLocality": "Varberg",
                "addressCountry": "Sweden"
            },
            "author": [
                {
                "@type": "CreativeWork",
                "name": "Some fundamental factors for obtaining sharp thermal curves",
                "publisher": "Transactions of the American Society for Steel Treating",
                "author": ["Karl Gustaf Lund", "C. Benedicks", "W. H. Dearden"],
                "datePublished": "1925"
                },
                {
                "@type": "CreativeWork",
                "name": "Contemporary production of saw blades, circular saws, and machine knives",
                "publisher": "Timber Industry",
                "author": "Karl Gustaf Lund",
                "datePublished": "1931"
                }
            ],
            "interests": "Hunting and Fishing"
    }
}

## Output

```{r}
library(ggtext)
library(showtext)
font_add_google("IBM Plex Mono", "ibm")
showtext_auto(enable = TRUE)

timeline <- tibble(
    event = c(
        "Birth", "Graduated from Bergsskolan in Filipstad",
        "Specialized studies at the Royal Institute of Technology (KTH)",
        "Studied at the Institute of Metallurgy and Stockholm University",
        "Marriage to Sigrid Johansson",
        "Birth of Ingvar", "Birth of Lennart",
        "Chemist at Strömsnäs Järnverks A-B in Degerfors",
        "Metallurgist and chemist at Westinghouse Electric & Manufacturing Co.",
        "Chief metallurgist at Laclede Steel Co. in Alton, Illinois, USA",
        "Furnace and steelworks engineer at A-B Iggesunds Bruk",
        "Site manager at Gunnebo Bruks Nya A-B, Varbergsverket",
        "Member of the municipal executive committee",
        "Deputy chairman of the economic department",
        "Deputy member of the board of the power plant",
        "Chairman of Varbergs Sparbank",
        "Employer representative in the district council of the county labor board",
        "Member of the board of Varbergs Luftskyddsförening",
        "Secretary of Varbergs Högerförening",
        "Chairman of the railway sick fund",
        "Plant Society for Small Bird Friends",
        "Traveled to Germany",
        "Traveled to Denmark and Czechoslovakia",
        "Traveled to Austria",
        "Traveled to the USA",
        "*Wrote 'Some fundamental factors for obtaining sharp thermal curves'",
        "*Wrote 'Contemporary production of saw blades, circular saws, and machine knives'",
        "Hobbies: hunting and fishing"
    ),
    start_year = c(
        1893, 1917, 1920, 1921, 1936, 1938, 1942, 1918, 1923, 1927, 1929,
        1931, NA, NA, NA, NA, NA, NA, NA, NA, NA, 1921, 1921, 1921, 1923, 1925, 1931, NA
    ),
    end_year = c(
        1893, 1917, 1922, 1922, 1936, 1938, 1942, 1920, 1929, 1927, 1931,
        NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 1923, 1923, 1921, 1929, NA, NA, NA
    )
)

timeline %>%
    filter(!is.na(start_year)) %>%
    arrange(start_year) %>%
    mutate(end_year_imputed = case_when(
        is.na(end_year) ~ lead(start_year),
        is.na(lead(start_year)) ~ start_year,
        TRUE ~ end_year
    )) %>%
    mutate(
        event = fct_reorder(event, start_year, .desc = TRUE)
    ) %>%
    # pivot_longer(c(start_year, end_year_imputed)) %>%
    # ggplot(aes(value, english_occupation, colour = name)) +
    ggplot(aes(xmin = start_year, xmax = end_year_imputed, y = event)) +
    geom_linerange(colour = "gray50", size = 2) +
    geom_point(aes(x = end_year_imputed), colour = "#be5d2b", size = 4) +
    geom_point(aes(x = start_year), colour = "#2B8CBE", size = 4) +
    labs(
        title = "Karl Lund's life course",
        subtitle = "Showing <span style = 'color:#2B8CBE;'>start dates</span> and <span style = 'color:#be5d2b;'>end dates</span> for his various roles and life events",
        x = NULL,
        y = NULL
    ) +
    theme(
        legend.position = "none",
        plot.subtitle = element_markdown(),
        plot.title.position = "plot",
        text = element_text(family = "ibm", size = 13)
    )

## Birthplaces mapped

Sample of birthplaces of individuals with surnames beginning with A in Götaland 1948:

```{r}
knitr::include_url("assets/sweden_map_checkboxes.html", height = "800")
```

## Birthplaces mapped

@@

## Why is classification important?

We often want to classify an observation into a group in order to make inference about the group.

```{r}
library(tidyverse)
library(gt)

skills_table <- tibble(
    green_skills = c(
        "Renewable Energy System Design",
        "Sustainable Urban Planning",
        "Environmental Impact Assessment",
        "Waste Management and Recycling",
        "Water Resource Management",
        "Sustainable Agriculture Practices",
        "Energy Efficiency Analysis",
        "Carbon Footprinting and Reduction",
        "Green Building Design",
        "Conservation Biology"
    ),
    brown_skills = c(
        "Petroleum Geology",
        "Oil and Gas Drilling Techniques",
        "Fossil Fuel Plant Operations",
        "Pipeline Design and Maintenance",
        "Refinery Operations",
        "Coal Mining Techniques",
        "Fossil Fuel Transportation Logistics",
        "Emission Control Technologies",
        "Fossil-Based Power Generation",
        "Risk Assessment in Fossil Fuel Industries"
    )
)

skills_table <- skills_table %>%
    gt() %>%
    tab_header(
        title = md("**Green and brown skills to be grouped**")
    ) %>%
    cols_label(
        green_skills = "Green Skills",
        brown_skills = "Brown Skills"
    ) %>%
    tab_style(
        style = cell_fill(color = "#b7e8b7"),
        locations = cells_body(columns = green_skills)
    ) %>%
    tab_style(
        style = cell_fill(color = "#e6b27e"),
        locations = cells_body(columns = brown_skills)
    )

skills_table %>%
    tab_options(table.font.size = 32)
```

## My use case: how to measure career progression

@@

- Here we do not say which skills are green or brown, we just use the `kmeans` algorithm to create two groups based on the embeddings:

![](assets/umap_1.png){fig-align="center" width=100%}

@@

# New terms to classify
new_terms = ["Sustainable textile recycling", "Coal burning technology"]

# Encode the new terms using the same model
new_embeddings = model.encode(new_terms)

# Transform the new embeddings using the already fit UMAP model
new_embeddings_umap = umap_model.transform(new_embeddings)

# Use the trained KMeans model to predict the cluster for the new embeddings
new_clusters = kmeans.predict(new_embeddings_umap)

# Mapping clusters to groups for readability
group_mapping = {1: 'Green Skills', 0: 'Brown Skills'}
classified_groups = [group_mapping[cluster] for cluster in new_clusters]

# Output the classification
for term, group in zip(new_terms, classified_groups):
    print(f"'{term}' is classified as: {group}")
```

- 'Sustainable textile recycling' is classified as: Green Skills

- 'Coal burning technology' is classified as: Brown Skills

## Visualize classification

![](assets/umap_2.png){fig-align="center" width=100%}

## Assign HISCOs: process

@@

## How to get a text embedding?

@@

## Assign HISCOs: example

@@

## Assign HISCOs: codes projected from 1536 dimensions to 2 with UMAP

@@

## Assign HISCOs: Watchmaker Apprentice

@@

## Assign HISCOs: Shipyard Assistant Engineer

@@

## Keep titles in Swedish?

@@

## SSYK and Swedish BERT examples:

@@

## Can we score how successful this is?

>>>>

    - It made individuals richer 💰🔋
    - It lowered inequality 📉🤝
    - It allowed structural change on the labor market 🔧💼
    - Middle-skill workers were drawn from the local area to the new jobs 🏙🏃‍♀️

@@

@@

@@

We provide the system a schema to structure the record in, e.g. keys.

We provide the system the Swedish text

We specify the JSON response format we want

@@

::: {.columns}
::: {.column}

```{python, eval=FALSE, echo=TRUE}
schema = {
    "type": "object",
    "required": [
        "full_name",
        "location",
        "occupation",
        "birth_details",
        "education",
        "career",
        "family",
    ],
    "properties": {
        "full_name": {"type": "string"},
        "location": {"type": "string"},
        "occupation": {"type": "string"},
        "birth_details": {
            "type": "object",
            "properties": {
                "date": {"type": "string"},
                "place": {"type": "string"},
                "parents": {"type": "string"},
            },
            "required": ["date", "place"],
        },}}"
```

:::
::: {.column}

@@

def structure_biography_info(page_text):
    try:
        # Create a prompt for the system
        structure_prompt = f"Task: read the schema and return RFC compliant JSON information about the Swedish individuals from the 1950 biographical dictionary 'Vem är Vem' that is provided below. Use a numeric index for each biography in your JSON output and return information about all of them, including all career information available. Keep the biographic descriptions in Swedish and remove any abbreviations based on your knowledge, e.g. 'fil. kand.' is 'filosofie kandidat', and 'Skarab. l.' is 'Skaraborgs Län'. Put years in full based on context. Put dates in dd/mm/yyyy format where possible. If there is no information for a key, leave it out. If there is no information for a required key, put NULL as the value.\nHere is the schema: {schema}.\nHere is the text: {page_text}. Go!"
        structure_response = client.chat.completions.create(
            model="gpt-3.5-turbo-1106",
            response_format={"type": "json_object"},
            messages=[
                {
                    "role": "system",
                    "content": "You are an expert on Swedish biographies.",
                },
                {"role": "user", "content": f"{structure_prompt}"},
            ],
        )

        structured_biography_info = json.loads(
            structure_response.choices[0].message.content
        )

        return structured_biography_info

    except Exception as e:
        print(f"Error in structure_biography_info: {e}")
        return None
```

## Add coordinates to firms and individuals

@@

## Two-part classification process

Makes use of occupational title, e.g. "Civilingenjör, teknisk fysik" and workplace information, e.g. "Strömsnäs Järnverks A-B".

@@

```{r}
#| label: fig-hisco-cluster
#| fig-cap: "Scatter plot showing the relative similarity of occupational titles in the HISCO schema"

knitr::include_graphics("assets/hisco_cluster.png")
```

## Use clusters to classify individuals into the groups

```{r}
#| label: fig-hisco-counts
#| fig-cap: "Most common HISCO occupations in Vem är Vem"
#| fig-width: 5
#| out-width: 5in

knitr::include_graphics("assets/hisco_counts.png")
```

@@

@@

## Cluster firms by location

```{r}
#| label: fig-map-common-types
#| fig-cap: "Map of the geographic clusters of businesses by most common business type"
#| fig-width: 5
#| out-width: 5in

knitr::include_graphics("assets/map_most_common_businesses.png")
```

# Appendix {#sec-appendix background-color="#1C9E77" visibility="uncounted"}

## How to get a text embedding?{visibility="uncounted"}

@@

## Assign HISCOs: example{visibility="uncounted"}

@@

## Assign HISCOs: codes projected from 1536 dimensions to 2 with UMAP{visibility="uncounted"}

@@

## Assign HISCOs: Watchmaker Apprentice{visibility="uncounted"}

@@

## Assign HISCOs: Shipyard Assistant Engineer{visibility="uncounted"}

@@

## Keep titles in Swedish?{visibility="uncounted"}

@@

## SSYK and Swedish BERT examples:{visibility="uncounted"}

@@

## Can we score how successful this is?{visibility="uncounted"}
